### 2024-06-05 journal/response.txt

> Reviewer 1
> Minor corrections:
> 
> line 080, right column. Emphasize that the number of CAS operations
> (log p) is another measure of the step complexity, different from
> the amortized step complexity.

Done.

> line 465. The notion sum_enq and sum_deq is confusing. Probably use
> count_enq and count_deq instead ? Or cnt_enq and cnt_deq ?

We were concerned that count might sound like a count of the number of enqueue or dequeue operations in a single block.  We instead used sum to remind the reader that the fields of Block i store the prefix sum of the number of enqueue or dequeue operations in in Blocks 1..i (as mentioned when the fields are introduced on page 11).

@@

> lines 594 - 608 (lines 24 - 37 in the code). Add a more detailed
> comments that explain the Refresh operation.

Done.  We have added many comments throughout the pseudocode.

> line 619 (line 44 in the code). It's not clear why the size of the
> queue is calculated correctly. Why if some of the dequeues that are
> counted by num_deq are empty and do not change the size of the
> queue ? Some intuition here would be helpful.

Some explanation has been added in the text (see magenta text on p.12)

@@

> lines 643 - 645 (lines 67 - 69 of the code). Add comments that help
> to understand the calculation of the dequeue's rank.

Done.

> line 658 (line 81 in the code). Add additional comment that
> doubling search is used here to find the range of the binary
> search.

Done.  We added to the text description of FindResponse (p.17) to explain that the doubling search first finds the left end of the range and then performs a binary search.

@@

> line 1304, right column. "...will also keep the running time of our
> operations polylogarithmic *on $p$ and $q$*"

Done.

> Reviewer 2
>
> Then, the paper presents a modification of the first algorithm that
> uses bounded memory, more specifically polynomial on the number of
> processes and the size of the queue in the linearization, and with
> polylogarithmic amortized step complexity. As one would naturally
> expect, this version adds a sort of garbage collector mechanism that
> removes blocks of memory that are not needed anymore. I found this
> part of the paper harder to understand. I did not understand the need
> of using red-black trees, which I think complicates everything. I
> have no reason to think the implementation is wrong, all looks
> convincing to me, but I wonder if there is a way to simplify the
> presentation.

For the space-bounded version, we replace an infinite array by a red-black tree (using the array index as the key).  This allows us to append elements, access elements, and split off the obsolete elements, all in logarithmic time.  We realize that this additional data structure complicates the algorithm, but we were unfortunately unable to come up with a simpler way to do it.  We hoped that presenting the simpler, space-unbounded version first would make it easier to understand the space-bounded version.  We did our best to explain the transformation from the arrays to the RBTs in the first page of Section 7.

@@

> Specific comments:
> 
> - ln 37, left. Missing blank space before ‘and’.

Fixed.

> - I find the related work section complete. However, I wonder if it is worth citing a few works about memory reclamation, due to your bounded memory implementation.

Although our space-bounded algorithm unlinks objects that are no longer needed from our data structure, we consider the problem of reclaiming that memory to be an orthogonal issue.  There is a rather large body of literature on memory reclamation for concurrent data structures, so we have added (at line 110, left column, of the introduction) a reference to a very recent paper that includes a large bibliography of work in this area, to help the reader find relevant literature on it.

@@

> - ln 235, left. Repeated ‘words’.

Fixed.

> - For completeness, I suggest to add a preliminaries section, with a reminder of the main definitions such as linearizability, step complexity, and so on.

A new section 3 on the model of computation has been added to include
these definitions.

@@

> - ln 1442. Why not Help(), in accordance to Dequeue(), for example?

We removed the parentheses from Dequeue for consistency.

> - Sec 7. Your algorithms are adaptive on the size of the queue but not on the number of processes. I wonder if an interesting future direction would be to make your algorithms fully adaptive.

Yes, this was mentioned in sentence 3 of the conclusion (now Section 8).

### 2024-06-05 journal/algorithm-GC.tex
working on reviewer 2's comments

The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index~0.
Any access to an entry of the \fld{blocks} array is replaced by a search in the RBT.
The node's \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead simply find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in constant time, without affecting the time of other RBT operations.
Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.

@@

\Function{Object}{Dequeue()}{} 
    \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}},$\\
	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}+1$,
	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}%
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{\rev{compute response \op{Dequeue} should return}}
\EndFunction{Dequeue}

>>>>

The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index~0.
Any access to an entry of the \fld{blocks} array is replaced by a search in the RBT.
The node's \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead simply find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in \rev{$O(1)$ steps}, without affecting the step complexity of other RBT operations.
Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.

@@

\Function{Object}{Dequeue}{} 
    \State \gc{\var{h} \assign\ \Call{MaxBlock}{\var{leaf}[\var{id}].\fld{blocks}}.\fld{index}+1}
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{enq}},$\\
	    \fld{sum\sub{deq}}\ \assign\ $\var{leaf}[\var{id}].\fld{blocks}[\gc{h-1}].\fld{sum\sub{deq}}+1$,
	    \gc{\fld{index} \assign $h$}}\label{deqNewGC}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \op{Dequeue} to \var{leaf}[\var{id}].\fld{blocks}, propagate to root}}%
    \gc{\State \Return \Call{CompleteDeq}{\var{leaf}[\var{id}], $h$}}\Comment{\rev{compute response \op{Dequeue} should return}}
\EndFunction{Dequeue}

### 2024-06-05 journal/boundedspace.tex

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
RBT make a new copy of each RBT node $x$ that must be modified, as well
as each RBT node along the path from the RBT's root to~$x$.
The RBT reachable from the new copy of the root is the result of applying the RBT operation.
This  adds only a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$\var{v}$ in the \ordering\ tree, 
it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.

To prevent RBTs from growing without bound, we must discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the running time of our operations polylogarithmic \rev{in $p$ and $q$}.
Blocks should be kept if they contain operations still in progress.
Moreover, a block containing an \op{Enqueue}($x$) operation must be kept until $x$ is dequeued.

To maintain good amortized time, we periodically do a garbage collection (GC) phase.
If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
To determine which blocks can be thrown away, we use a global array $\var{last}[1..p]$ where 
each process writes the index of the last block in the root
containing a null \dequeue\ or an \enqueue\ whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending \dequeue\ by 
computing the \dequeue's response and writing it in the block in the leaf that represents the \dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\fld{root.blocks}[1..m-1]$ have all been dequeued, so GC can discard all subblocks of those.
Fortunately, there is an  RBT \op{Split} operation that can remove
these obsolete blocks from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.

>>>>

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a block, just as the \op{CAS} on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful \op{CAS}.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other block in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all RBT nodes are immutable, and operations on the 
RBT make a new copy of each RBT node $x$ that must be modified, as well
as each RBT node along the path from the RBT's root to~$x$.
The RBT reachable from the new copy of the root is the result of applying the RBT operation.
This \rev{increases the number of steps taken by any routine designed for a (sequential) RBT only by a constant factor, assuming that nodes are accessed only by following pointers from the root (as is the case for all RBT operations we require)}.
Once a process has performed an update to the RBT representing the blocks of a node 
$\var{v}$ in the \ordering\ tree, 
it uses a CAS to swing $\var{v}$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent RBTs in a similar way for a snapshot data structure.

To prevent RBTs from growing without bound, we must discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the \rev{step complexity} of our operations polylogarithmic \rev{in $p$ and $q$}.
Blocks should be kept if they contain operations still in progress.
Moreover, a block containing an \op{Enqueue}($x$) operation must be kept until $x$ is dequeued.

To maintain good amortized \rev{step complexity}, we periodically do a garbage collection (GC) phase.
If a \op{Refresh} on a node adds a block whose \var{index} is a multiple of $G=p^2\ceil{\log p}$, it does GC to remove obsolete blocks from the node's RBT.
To determine which blocks can be thrown away, we use a global array $\var{last}[1..p]$ where 
each process writes the index of the last block in the root
containing a null \dequeue\ or an \enqueue\ whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending \dequeue\ by 
computing the \dequeue's response and writing it in the block in the leaf that represents the \dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\fld{root.blocks}[1..m-1]$ have all been dequeued, so GC can discard all subblocks of those.
Fortunately, there is an  RBT \op{Split} operation that can remove
these obsolete blocks from an RBT using \rev{a logarithmic number of steps} \cite[Sec.~4.2]{Tar83}.

### 2024-06-05 journal/correctness.tex

The following two lemmas show that no operation appears in more than one block of the root.
\begin{lemma} \label{lem::headProgress}
 If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$, then
 \begin{eqnarray*}
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{left}} \mbox{ and}\\
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{right}}.
 \end{eqnarray*}
\end{lemma}
\begin{proof}
Let $B$ be the block in $\var{v}.\fld{blocks}[b]$.
Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
read $b$ from $\var{v}.\head$ at line \ref{readHead}.
At that time, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
created $B$ started.
It follows from Observation \ref{nonDecreasingHead} that the value that 
line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b$) 
stores in $B.\fld{end\sub{left}}$.
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
%of these calls to \op{CreateBlock} 
satisfy the claim.
\end{proof}

@@

We show that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the concurrent execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  We also show each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would in the sequential execution~$L$ \rev{(i.e., the sequential execution of the operations in the order given by $L$)}.

@@

Next, we show each operation returns the same response as it would in the sequential execution~$L$.

\begin{lemma}\label{linearCorrect}
Each terminating \dequeue\ returns the response it would in the sequential execution~$L$.
\end{lemma}
\begin{proof}
If a \dequeue\ $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th \dequeue\ in the block 
$B=\var{root}.\fld{blocks}[b]$.
$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of \enqueues\ in $B$.
Since the \enqueues\ in block $B$ precede the \dequeues,
the queue is empty when the $i$th \dequeue\ of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ \dequeues\ in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
Thus, in the sequential execution~$L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required \enqueue.
\end{proof}

@@

>>>>

The following two lemmas show that no operation appears in more than one block of the root.
\begin{lemma} \label{lem::headProgress}
 If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$, then
 \begin{eqnarray*}
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{left}} \mbox{ and}\\
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{right}}.
 \end{eqnarray*}
\end{lemma}
\begin{proof}
Let $B$ be the block in $\var{v}.\fld{blocks}[b]$.
Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
read $b$ from $\var{v}.\head$ at line \ref{readHead}.
At that \rev{read}, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
created $B$ started.
It follows from Observation \ref{nonDecreasingHead} that the value that 
line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b$) 
stores in $B.\fld{end\sub{left}}$.
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
%of these calls to \op{CreateBlock} 
satisfy the claim.
\end{proof}

@@

We \rev{next show in Lemma \ref{linearSat}} that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the concurrent execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  \rev{Then, we show in Lemma \ref{linearCorrect}} that each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would \rev{if the operations were performed sequentially in the order given by $L$}.

@@

Next, we show each operation returns the same response as it would \rev{if the operations were performed sequentially in the order $L$}.

\begin{lemma}\label{linearCorrect}
Each terminating \dequeue\ returns the response it would \rev{if the operations
were performed sequentially in the order given by the linearization $L$}.
\end{lemma}
\begin{proof}
If a \dequeue\ $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th \dequeue\ in the block 
$B=\var{root}.\fld{blocks}[b]$.
$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of \enqueues\ in $B$.
Since the \enqueues\ in block $B$ precede the \dequeues,
the queue is empty when the $i$th \dequeue\ of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution \rev{defined by} $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution \rev{defined by} $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ \dequeues\ in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
Thus, in the sequential execution \rev{defined by} $L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required \enqueue.
\end{proof}

@@

\rev{We remark that, if execution $\alpha$ is a prefix of another execution $\alpha'$,
then the linearization associated with $\alpha$ is a prefix of the linearization
assocated with $\alpha'$, so our implementation is also strongly linearizable,
according to the definition in \cite{GHW11}.}
